---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

Name: Mayuri Patel

Email: mpate131@uncc.edu

title: "Lab7_AdvStats"


(1)	We again return to our RNA seq dataset of E. Coli genes from mice.  
The URL is here:
http://afodor.github.io/classes/stats2015/longitdunalRNASeqData.zip
As before, read and normalize the counts table ( “nc101_scaff_dataCounts.txt “ into R).  For example:
setwd("C:\\somewhere")

myT<-read.table("nc101_scaff_dataCounts.txt",sep="\t",header=TRUE,row.names=1)
## remove rare genes
myT <- myT[ apply( myT,1, median)> 5,]

myTNorm <- myT
for ( i in 1:ncol(myT))
{
	colSum = sum(myT[,i])
	myTNorm[,i] =myTNorm[,i]/colSum
}
(The first 3 columns are “day 2”, the next 3 columns are “week 12” and the last 5 are “week 18”).  

- Reading the file of datacount from the directory:

```{r}
#set the working directory to the file location
setwd("C:/Users/mpate131/Desktop/AdvStats")

#read the table
myTable<-read.table("nc101_scaff_dataCounts.txt",sep="\t",header=TRUE,row.names=1)
head(myTable) #viewing the data using head function
```

- Normalize the data: 

```{r}
# remove rare genes
myTable <- myTable[ apply( myTable,1, median)> 5,]

myTNorm <- myTable
#For loop each column
for ( i in 1:ncol(myTable))
{
	colSum = sum(myTable[,i]) # all rows and each column and summing all the column
	myTNorm[,i] =myTNorm[,i]/colSum # normalizing the data and saving it in the empty variable myTNorm.

}
head(myTNorm) #viewing the data using head function
```


(A)	For each row in the spreadsheet, perform a one-way ANOVA with categories “day 2”, “week 12” and “week 18”.  Plot out the histogram of all p-values.  How many genes are significant at a BH FDR-corrected 0.05 threshold. 


```{r}

#replicating the categories of the sample here.
categories_value <- c(rep("D2",3), rep("W12",3), rep("W20",5))

#empty variable to save the p values
anova_pVal <- c()

#get each row of data and cast it to numeric
for( i in 1:nrow(myTNorm)){
	myData <- as.numeric(myTNorm[i,] ) 
  
	##  building linear models with myData as the y-variable and categories as x variable
	mylinearmodel <- lm(myData ~ categories_value, x=TRUE)
	anova_pVal <- c(anova_pVal, anova(mylinearmodel)$ "Pr(>F)"[1]) # extracting the p value from the linear model
}
# histogram of the p-values.
hist(anova_pVal,xlab = "P-values", ylab ="Frequency",main ="Histogram of p-values for one-way ANOVA with categories") 

#For p <0.05 BH FDR corrected:
anova_BH <- p.adjust(anova_pVal, method="BH") # using method BH
anova_BHpval <- sum(anova_BH<0.05) # summing all the p values <0.05
print(paste("The number of genes significant at p-values <0.05 BH FDR corrected for one-way ANOVA with categories “day 2”, “week 12” and “week 18” is :", anova_BHpval))

```


(B)	Next make an ANOVA as a linear regression as a function of time (so 2 days, 86 days and 128 days).  Plot out the histogram of all p-values.  How many genes are significant at a BH FDR-corrected 0.05 threshold. 

```{r}
#replicating the time of the sample here.
time_value <- c(rep(2,3), rep(86,3), rep(128,5))

##empty variable to save the p values
aovLR_pVal <- c()


#get each row of data and cast it to numeric
for( i in 1:nrow(myTNorm)){
	myData <- as.numeric(myTNorm[i,] ) 
  
	##  building linear models with myData as the y-variable and time as x-variable
	mylinearmodel1 <- lm(myData ~ time_value, x=TRUE)
	aovLR_pVal <- c(aovLR_pVal, anova(mylinearmodel1)$ "Pr(>F)"[1]) # extracting the p value from the linear model
}
#histogram of the p-values.
hist(aovLR_pVal,xlab = "P-values", ylab ="Frequency",main ="Histogram of p-values for ANOVA as linear regression")

#For p <0.05 BH FDR corrected:
aovLR_BH <- p.adjust(aovLR_pVal, method="BH") # using method BH
aovLR_BHpval <- sum(aovLR_BH <0.05) # summing all the p values <0.05
print(paste("The number of genes significant at p-values <0.05 BH FDR corrected for ANOVA as linear regression for time 2 days, 86 days and 128 days is :", aovLR_BHpval))
```


(C)	Finally, for each row in the spreadsheet perform an ANOVA comparing the three-parameter model from (A) and the two parameter model from (B).  Plot out the histogram of all p-values.  For how many genes is there a significant difference between these two models at a BH FDR-corrected threshold. 


```{r}
#empty variable to save the p values
pvalue_compare <- c()

#get each row of data and cast it to numeric
for( i in 1:nrow(myTNorm)){
	myData <- as.numeric(myTNorm[i,] ) 
	
  anov_3para <- lm(myData ~ categories_value,x=TRUE) # three-parameter model
  
  anov_2para <- lm(myData ~ time_value, x=TRUE) # two-parameter model
  
  fullmodel <- sum(residuals(anov_3para)^2) # calculating residual sum for 3 parameter model
  reducemodel <- sum(residuals(anov_2para)^2) # calculating residual sum for 2 parameter model
  
  #calculating the F value using reduce and full model along with df (n=sample size=> 11-3/11-2)
  F = ((reducemodel - fullmodel)/(1))/(fullmodel/8)
  
  #Now calculating the p value using pf function along with df1 and df2 
  pval <- pf(F, 1,8,lower.tail = FALSE)
  
  pvalue_compare <- c(pvalue_compare, pval)# saving all p values into an empty variable.
  
}
hist(pvalue_compare,xlab = "P-values", ylab ="Frequency",main ="Histogram of p-values for ANOVA comparing 3 and 2-parameter model")


#For p <0.05 BH FDR corrected:
pval_BH <- p.adjust(pvalue_compare, method="BH") 
BHpval <- sum(pval_BH <0.05)
print(paste("The number of genes significantly different between three-parameter and two-parameter model at p-values <0.05 BH FDR corrected is :", BHpval))
```



(D)	 Make three graphs showing the relative abundance of the most significant gene under each of the three ANOVA models.  For (A) and (C), the x-axis will the category (day 3, week 12 and week 18) and the y-axis will be the relative abundance.  Be sure to properly label and title all graphs and axes.  For (B) the x-axis will be time (in days) and the y-axis will be the relative abundance. For the graph of the top hit from (B), include the regression line for the plot from (B).


```{r}
#empty variables to save the p values from each of the three models.
pValuesOneWayAnova <- vector()
pValuesRegression <- vector()
pValueModelDiff <- vector()

#index of the rows
index <- vector()

categories_value <- c(rep("D2",3), rep("W12",3), rep("W20",5))

for( i in 1:nrow(myTNorm)){
  
	index[i] <- i # indexing the rows
	
	#get each row of data and cast it to numeric
	myData <- as.numeric(myTNorm[i,] ) 
  
	#---model 1----
	#  building linear models with myData as the y-variable 
	mylinearmodel <- lm(myData ~ categories_value, x=TRUE)
	pValuesOneWayAnova <- c(pValuesOneWayAnova, anova(mylinearmodel)$ "Pr(>F)"[1])# extracting the p value from the linear model one-way anova
	
	#---model 2----
	##  building linear models with myData as the y-variable 
	mylinearmodel1 <- lm(myData ~ time_value, x=TRUE)
	pValuesRegression <- c(pValuesRegression, anova(mylinearmodel1)$ "Pr(>F)"[1])# extracting the p value from the linear model anova as linear regression
	
  #---model 3----
  fullmodel <- sum(residuals(mylinearmodel)^2)# calculating residual sum for 3 parameter model
  
  reducemodel <- sum(residuals(mylinearmodel1)^2)  # calculating residual sum for 2 parameter model
  
  #calculating the F value using reduce and full model along with df (n=sample size=> 11-3/11-2)
  F = ((reducemodel - fullmodel)/(1))/(fullmodel/8)
  
  #Now calculating the p value using pf function along with df1 and df2  
  pval <- pf(F, 1,8,lower.tail = FALSE)
  pValueModelDiff <- c(pValueModelDiff, pval)# saving all p values into an empty variable.
	
}
#built a dataframe of p values from three models with the index number
myFrame <- data.frame( index, pValuesOneWayAnova,pValuesRegression,pValueModelDiff)
head(myFrame)
```

- (A) boxplot for one way ANOVA:

```{r}
# Here, we are ordering the p values of one way ANOVA:
myFrame <- myFrame[ order(myFrame$pValuesOneWayAnova),] 

# in this boxplot, we are taking the sample values of a most significant gene based on the ordered index number on x axis and y being the abundance of the gene
boxplot( as.numeric( myTNorm[ myFrame$index[1],]) ~ categories_value,xlab = "categories", ylab ="Relative abundance of NC101_03447 gene", main="Boxplot of ANOVA showing relative abundance of most significant gene",col = c("#00AFBB", "#E7B800", "#FC4E07") ) 
```

- (B) plot with regression line:

```{r}
# Here, we are ordering the p values of one way ANOVA as linear regression:
myFrame <- myFrame[ order(myFrame$pValuesRegression), ] 

#In this plot, we are taking the sample values of a most significant gene based on the ordered index number on x axis and y being the abundance of the gene
plot(time_value,as.numeric( myTNorm[ myFrame$index[1],]),xlab = "time(in days)", ylab ="Relative abundance of NC101_03467 gene",main="Plot with regression line for most significant gene using ANOVA LR",col="red",pch = 15)
# calculating the linear model to add the regression line
mylm <- lm(as.numeric( myTNorm[ myFrame$index[1],])~time_value)
abline(mylm, col="green") # This adds the regression line to the plot
```

- (C) Boxplot for model difference:

```{r}
# Here, we are ordering the p values of model difference:
myFrame <- myFrame[ order(myFrame$pValueModelDiff), ]

# In this boxplot, we are taking the sample values of a most significant gene based on the ordered index number on x axis and y being the abundance of the gene
boxplot( as.numeric( myTNorm[ myFrame$index[1],]) ~ categories_value,xlab = "categories", ylab ="Relative abundance of NC101_03447 gene", main ="Boxplot of model difference for most significant gene",col = c("#01AFBB", "#E7B810", "#FC4E08")  ) 
```

(E)	 Overall, do you think the three parameter model in (A) or the two-parameter model in (B) is more appropriate for these data?  Justify your answer.
- From my perspective, three parameter model is more appropriate for these data because in the full model/ three parameter we assume that each observation is a function of a specific group mean, so there are more parameters (one for each group's mean--hence "full").The full model appears to describe the trend in the data better than the reduced model. Whereas in the reduced model we assume that each observation is a function of the grand mean, so there are fewer parameters (hence, "reduced") to compare. A reduced model does not include all the possible terms to be used for these data.The 2 parameter model makes assumptions of linearity that are not appropriate for a substantial subset of the genes. 
